Spatial Reasoning and Data Displays

### 2015-06-16 Paper/InfoVis2015Revision/TestingVisualAptitude.tex
Addressing reviewer comments

\begin{table*}[htb]\centering
\begin{tabular}{rlllc}
\hline
  & Card Rotation & Paper Folding & Figure Classification & Visual Search  \\\hline
ISU Students & 83.4 (24.1) 
             & 12.4 (3.7)
             & 57.0 (23.8)\footnotemark[1]
             & 21.9 (2.3)\\
Scaled Scores & 88.0 (34.8)
              & 13.8 (4.5)
              & 58.7 (14.4)\footnotemark[2]
              & -- \\
Unscaled Scores & 44.0 (24.6)\footnotemark[3]
                & 13.8 (4.5)
                & \parbox[c]{.2\linewidth}{M:~120.0~(30.0)\\ F:~114.9~(27.8)}
                & --\\\hline
{\footnotesize Population}    
              & \parbox[t]{.15\linewidth}{\footnotesize approx.\ 550 male\\ naval recruits} 
              & \parbox[t]{.17\linewidth}{\footnotesize 46 college students\\(1963~version)}
              & \parbox[t]{.2\linewidth}{\footnotesize suburban 11th \& 12th \\ grade students\\(288-300 M, 317-329 F)}
              & \\\hline
\end{tabular}
\caption[Comparison of scores for cognitive tasks.]{Comparison of scores from Iowa State students and scores reported in~\protect\cite{ekstrom1976manual}. Scaled scores are calculated based on information reported in the manual, scaled to account for differences in the number of questions answered during this experiment. Data shown are from the population most similar to ISU students, out of the data available. The visual search task~\protect\cite{goldstein1973validity,demita1981validity,moerland1986neuropsychological} is not part of the Kit of Factor Referenced Cognitive Test data, and thus we do not have comparison data for the form used in this experiment.
\label{tab:scorecomparison}}
\end{table*}

@@

% latex table generated in R 3.2.0 by xtable 1.7-1 package
% Mon Jun 15 15:31:31 2015
\begin{table}[htb]
\centering
\caption{Importance of principal components in an analysis of four tests of spatial ability: figure classification, paper folding, card rotation, and visual search.\label{tab:PCAvariance4}} 
{\footnotesize
\begin{tabular}{rrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 \\ 
  \hline
Standard deviation & 1.61 & 0.81 & 0.73 & 0.49 \\ 
  Proportion of Variance & 0.64 & 0.16 & 0.13 & 0.06 \\ 
  Cumulative Proportion & 0.64 & 0.81 & 0.94 & 1.00 \\ 
   \hline
\end{tabular}
}
\end{table}

@@

% latex table generated in R 3.2.0 by xtable 1.7-1 package
% Mon Jun 15 15:31:31 2015
\begin{table}[htb]
\centering
\caption{Rotation matrix for principal component analysis of the four cognitive tests (visual search, paper folding, card rotation, figure classification).\label{tab:PCArotation4}} 
\begin{tabular}{rrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 \\ 
  \hline
card.rot & 0.55 & -0.19 & -0.38 & 0.72 \\ 
  fig.class & 0.46 & 0.58 & 0.66 & 0.14 \\ 
  folding & 0.52 & 0.33 & -0.53 & -0.59 \\ 
  vis.search & 0.46 & -0.72 & 0.38 & -0.34 \\ 
   \hline
\end{tabular}
\end{table}

@@

Next, we add the lineup results to our principal components analysis. The results of the principal component analysis of the four cognitive tests and lineup results are similar to the results of the cognitive tests alone; this suggests that lineups require at least some of the skills measured by the other cognitive tests. Table~\ref{tab:PCAvariance5} shows the importance of each principal component; from the distribution of the variance components, we see that the lineup test spans an additional dimension within the space of the four established tests. 
% latex table generated in R 3.2.0 by xtable 1.7-1 package
% Mon Jun 15 15:31:31 2015
\begin{table}[ht]
\centering
\caption{Importance of principal components, analyzing all five tests.\label{tab:PCAvariance5}} 
{\footnotesize
\begin{tabular}{rrrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 & PC5 \\ 
  \hline
Standard deviation & 1.73 & 0.84 & 0.75 & 0.70 & 0.48 \\ 
  Proportion of Variance & 0.60 & 0.14 & 0.11 & 0.10 & 0.05 \\ 
  Cumulative Proportion & 0.60 & 0.74 & 0.85 & 0.95 & 1.00 \\ 
   \hline
\end{tabular}
}
\end{table}

% latex table generated in R 3.2.0 by xtable 1.7-1 package
% Mon Jun 15 15:31:31 2015
\begin{table}[ht]
\centering
\caption{PCA Rotation matrix for all five tests. The first principal component is essentially an average of all five tests. 
             \label{tab:PCArotation5}} 
\begin{tabular}{rrrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 & PC5 \\ 
  \hline
lineup & 0.42 & 0.49 & -0.46 & 0.60 & -0.10 \\ 
  card.rot & 0.50 & -0.30 & 0.28 & 0.23 & 0.73 \\ 
  fig.class & 0.43 & 0.45 & -0.15 & -0.75 & 0.18 \\ 
  folding & 0.47 & 0.07 & 0.68 & 0.04 & -0.56 \\ 
  vis.search & 0.41 & -0.69 & -0.48 & -0.15 & -0.33 \\ 
   \hline
\end{tabular}
\end{table}

@@

\end{knitrout}
Biplots \cite{biplots:gabriel} of the remaining components are provided in figure~\ref{fig:biplots5}. Figure classification is strongly related to lineups (PC2, PC3). Performance on the visual search task  is also related to lineup performance (PC3). These two components highlight the shared demands of the lineup task and the figure classification task: participants must establish categories from provided stimuli and then classify the stimuli accordingly.

@@

% latex table generated in R 3.2.0 by xtable 1.7-1 package
% Mon Jun 15 15:31:32 2015
\begin{table}[ht]
\centering
\caption{Estimates and significances of a linear model of lineup scores.} 
\label{tab:m1}
\begin{tabular}{rrrrr}
  \hline
 & Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
  \hline
(Intercept) & 14.1192 & 1.9149 & 7.37 & 0.0000 \\ 
  PC1 & 1.7672 & 0.5230 & 3.38 & 0.0018 \\ 
  MATH & 2.1246 & 0.8732 & 2.43 & 0.0202 \\ 
   \hline
\end{tabular}
\end{table}

%\afterpage{\clearpage}
%
\subsection{Lineup Types}
Each of the three sets of 20 lineups was taken from previous experiments examining which type of plot most effectively conveyed important characteristics of the underlying data set. 
\begin{figure*}[htbp]\centering
\subfloat[Set 1: Boxplot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups1/boxplot_1_15_5_4_13}}
\subfloat[Set 1: Density plot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups1/density_1_15_5_4_13}}
\subfloat[Set 1: Dotplot ]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups1/dotplot_1_15_5_4_13}}
\subfloat[Set 1: Histogram]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups1/histogram_1_15_5_4_13}}\\
\subfloat[Set 2: Boxplot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups2/file147f94f6965d7}}
\subfloat[Set 2: Boxplot with jittered points]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups2/file147f9e65f0e4}}
\subfloat[Set 2: Bee swarm boxplot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups2/file147f9fc58d73}}
\subfloat[Set 2: Violin plot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups2/file147f93acbb7e4}}\\
\subfloat[Set 3: QQ-plot with guide line]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups3/file13876a225d34-single}}
\subfloat[Set 3: QQ-plot with acceptance bands]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups3/file13877c93dc78-single}}
\subfloat[Set 3: QQ-plot rotated 45 degrees.]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups3/file13878e5c92f-single}}
\caption{Lineup plot types used in the experiment. \label{fig:lineups}}
\end{figure*}

\begin{figure}[ht]
\centering
\includegraphics[width=\linewidth]{Figure/fig-LineupTypeSPM-1}
\caption{Pairwise scatterplots of test scores, with lineups separated into the three lineup tasks. All lineup tasks are moderately correlated with the figure classification task, and while tasks 1 and 2 are most strongly correlated with figure classification, lineup task 3 is most strongly correlated with the card rotation and paper folding tasks. \label{fig:lineuptaskmatrix} \newtext{While the correlations shown in this graph are all less than 0.60, there are some surprisingly strong relationships for human subjects data.}}
\end{figure}
Figure~\ref{fig:lineuptaskmatrix} shows the correlations between the three lineup tasks and the figure classification, card rotation, and paper folding tasks. 
%The visual search task is only slightly correlated with the three lineup tasks and is therefore omitted from this figure.
Performance on lineup tasks 1 and 2, which dealt with the distributuion of two groups of numerical data, is most strongly correlated with the performance on the figure classification task, which measures general reasoning ability.
Performance on lineup task 3, which investigated the potential to visually identify nonnormality in residual QQ-plots, is more associated with the card rotation and paper folding tests, which measure visuospatial ability. This suggests that certain lineup tasks may require more visual ability than others; in the case of the QQ-lineups a successful evaluation  needed  participants to mentally rotate plots to compare vertical distances, requiring more mental manipulation than the first two lineup tasks.

>>>>

\begin{table*}[htb]\centering
\begin{tabular}{rlllc}
\hline
  & Card Rotation & Paper Folding & Figure Classification & Visual Search  \\\hline
ISU Students & 83.4 (24.1) 
             & 12.4 (3.7)
             & 57.0 (23.8)\footnotemark[1]
             & 21.9 (2.3)\\
Scaled Scores & 88.0 (34.8)
              & 13.8 (4.5)
              & 58.7 (14.4)\footnotemark[2]
              & -- \\
Unscaled Scores & 44.0 (24.6)\footnotemark[3]
                & 13.8 (4.5)
                & \parbox[c]{.2\linewidth}{M:~120.0~(30.0)\\ F:~114.9~(27.8)}
                & --\\\hline
{\footnotesize Population}    
              & \parbox[t]{.15\linewidth}{\footnotesize approx.\ 550 male\\ naval recruits} 
              & \parbox[t]{.17\linewidth}{\footnotesize 46 college students\\(1963~version)}
              & \parbox[t]{.2\linewidth}{\footnotesize suburban 11th \& 12th \\ grade students\\(288-300 M, 317-329 F)}
              & \\\hline
\end{tabular}
\caption[Comparison of scores for cognitive tasks.]{Comparison of scores from Iowa State students and scores reported in~\protect~\cite{ekstrom1976manual}. Scaled scores are calculated based on information reported in the manual, scaled to account for differences in the number of questions answered during this experiment. Data shown are from the population most similar to ISU students, out of the data available. The visual search task~\protect~\cite{goldstein1973validity,demita1981validity,moerland1986neuropsychological} is not part of the Kit of Factor Referenced Cognitive Test data, and thus we do not have comparison data for the form used in this experiment.
\label{tab:scorecomparison}}
\end{table*}

@@

% latex table generated in R 3.2.0 by xtable 1.7-1 package
% Tue Jun 16 12:39:22 2015
\begin{table}[htb]
\centering
\caption{Importance of principal components in an analysis of four tests of spatial ability: figure classification, paper folding, card rotation, and visual search.\label{tab:PCAvariance4}} 
{\footnotesize
\begin{tabular}{rrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 \\ 
  \hline
Standard deviation & 1.61 & 0.81 & 0.73 & 0.49 \\ 
  Proportion of Variance & 0.64 & 0.16 & 0.13 & 0.06 \\ 
  Cumulative Proportion & 0.64 & 0.81 & 0.94 & 1.00 \\ 
   \hline
\end{tabular}
}
\end{table}

@@

% latex table generated in R 3.2.0 by xtable 1.7-1 package
% Tue Jun 16 12:39:22 2015
\begin{table}[htb]
\centering
\caption{Rotation matrix for principal component analysis of the four cognitive tests (visual search, paper folding, card rotation, figure classification).\label{tab:PCArotation4}} 
\begin{tabular}{rrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 \\ 
  \hline
card.rot & 0.55 & -0.19 & -0.38 & 0.72 \\ 
  fig.class & 0.46 & 0.58 & 0.66 & 0.14 \\ 
  folding & 0.52 & 0.33 & -0.53 & -0.59 \\ 
  vis.search & 0.46 & -0.72 & 0.38 & -0.34 \\ 
   \hline
\end{tabular}
\end{table}

@@

Next, we add the lineup results to our principal components analysis. The results of the principal component analysis of the four cognitive tests and lineup results are similar to the results of the cognitive tests alone; this suggests that lineups require at least some of the skills measured by the other cognitive tests. Table~\ref{tab:PCAvariance5} shows the importance of each principal component; from the distribution of the variance components, we see that the lineup test spans an additional dimension within the space of the four established tests. 
% latex table generated in R 3.2.0 by xtable 1.7-1 package
% Tue Jun 16 12:39:22 2015
\begin{table}[ht]
\centering
\caption{Importance of principal components, analyzing all five tests.\label{tab:PCAvariance5}} 
{\footnotesize
\begin{tabular}{rrrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 & PC5 \\ 
  \hline
Standard deviation & 1.73 & 0.84 & 0.75 & 0.70 & 0.48 \\ 
  Proportion of Variance & 0.60 & 0.14 & 0.11 & 0.10 & 0.05 \\ 
  Cumulative Proportion & 0.60 & 0.74 & 0.85 & 0.95 & 1.00 \\ 
   \hline
\end{tabular}
}
\end{table}

% latex table generated in R 3.2.0 by xtable 1.7-1 package
% Tue Jun 16 12:39:22 2015
\begin{table}[ht]
\centering
\caption{PCA Rotation matrix for all five tests. The first principal component is essentially an average of all five tests. 
             \label{tab:PCArotation5}} 
\begin{tabular}{rrrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 & PC5 \\ 
  \hline
lineup & 0.42 & 0.49 & -0.46 & 0.60 & -0.10 \\ 
  card.rot & 0.50 & -0.30 & 0.28 & 0.23 & 0.73 \\ 
  fig.class & 0.43 & 0.45 & -0.15 & -0.75 & 0.18 \\ 
  folding & 0.47 & 0.07 & 0.68 & 0.04 & -0.56 \\ 
  vis.search & 0.41 & -0.69 & -0.48 & -0.15 & -0.33 \\ 
   \hline
\end{tabular}
\end{table}

@@

\end{knitrout}
Biplots~\cite{biplots:gabriel} of the remaining components are provided in figure~\ref{fig:biplots5}. Figure classification is strongly related to lineups (PC2, PC3). Performance on the visual search task  is also related to lineup performance (PC3). These two components highlight the shared demands of the lineup task and the figure classification task: participants must establish categories from provided stimuli and then classify the stimuli accordingly.

@@

% latex table generated in R 3.2.0 by xtable 1.7-1 package
% Tue Jun 16 12:39:23 2015
\begin{table}[ht]
\centering
\caption{Estimates and significances of a linear model of lineup scores.} 
\label{tab:m1}
\begin{tabular}{rrrrr}
  \hline
 & Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
  \hline
(Intercept) & 14.1192 & 1.9149 & 7.37 & 0.0000 \\ 
  PC1 & 1.7672 & 0.5230 & 3.38 & 0.0018 \\ 
  MATH & 2.1246 & 0.8732 & 2.43 & 0.0202 \\ 
   \hline
\end{tabular}
\end{table}

%\afterpage{\clearpage}
%
\subsection{Lineup Types}
Figure~\ref{fig:lineuptaskmatrix} shows the correlations between the three lineup tasks and the figure classification, card rotation, and paper folding tasks. 
%The visual search task is only slightly correlated with the three lineup tasks and is therefore omitted from this figure.
Performance on lineup tasks 1 and 2, which dealt with the distributuion of two groups of numerical data, is most strongly correlated with the performance on the figure classification task, which measures general reasoning ability.
Performance on lineup task 3, which investigated the potential to visually identify nonnormality in residual QQ-plots, is more associated with the card rotation and paper folding tests, which measure visuospatial ability. This suggests that certain lineup tasks may require more visual ability than others; in the case of the QQ-lineups a successful evaluation  needed  participants to mentally rotate plots to compare vertical distances, requiring more mental manipulation than the first two lineup tasks.

\begin{figure}[ht]
\centering
\includegraphics[width=\linewidth]{Figure/fig-LineupTypeSPM-1}
\caption{Pairwise scatterplots of test scores, with lineups separated into the three lineup tasks. All lineup tasks are moderately correlated with the figure classification task, and while tasks 1 and 2 are most strongly correlated with figure classification, lineup task 3 is most strongly correlated with the card rotation and paper folding tasks. \label{fig:lineuptaskmatrix} \newtext{While the correlations shown in this graph are all less than 0.60, there are some surprisingly strong relationships for human subjects data.}}
\end{figure}
